---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::knit_engines$set(sql = function(options) {
  code <- paste(options$code, collapse = "\n")
  extension_path <- normalizePath("build/release/ducknng.duckdb_extension", mustWork = FALSE)
  extension_path <- gsub("'", "''", extension_path, fixed = TRUE)
  code <- paste(sprintf("LOAD '%s';", extension_path), code, sep = "\n")
  if (isTRUE(options$eval)) {
    out <- system2("duckdb", "-unsigned", input = code, stdout = TRUE, stderr = TRUE)
    status <- attr(out, "status")
    if (!is.null(status) && status != 0) stop(paste(out, collapse = "\n"))
    knitr::engine_output(options, code, out)
  } else {
    knitr::engine_output(options, code, NULL)
  }
})
```

# ducknng

`ducknng` is a pure C DuckDB extension for DuckDB + [NNG](https://nng.nanomsg.org/) REQ/REP interop.

It gives DuckDB SQL a `nanonext`-like messaging surface and uses a thin, versioned RPC envelope with Arrow IPC payloads. The SQL ergonomics are inspired by [`r-lib/nanonext`](https://github.com/r-lib/nanonext), and the thin-envelope + Arrow IPC direction is informed by projects such as [`sounkou-bioinfo/mangoro`](https://github.com/sounkou-bioinfo/mangoro).

Today the extension already includes:

- server lifecycle control through `ducknng_start_server()`, `ducknng_stop_server()`, and `ducknng_list_servers()`
- generic NNG socket handles for `bus`, `pair`, `poly`, `push`, `pull`, `pub`, `sub`, `req`, `rep`, `surveyor`, and `respondent`
- raw socket send/recv primitives plus req-style raw request/reply helpers
- raw aio helpers for socket send/recv and req-style request/reply futures
- a low-level HTTP/HTTPS client helper for transport-adapter work
- explicit query-session helpers and TLS config handles

Transport is selected from the URL scheme. Today the NNG-facing operations work across `inproc://`, `ipc://`, `tcp://`, and `tls+tcp://`, and `ducknng_ncurl(...)` provides the first low-level HTTP/HTTPS client slice. The generic socket layer is explicitly modeled on `nanonext`: open a protocol-specific socket, dial or listen it, and then use raw send/recv or aio helpers as appropriate. The higher-level RPC helpers wrap manifest-declared request/reply methods, and the session helpers wrap the fixed `query_open` / `fetch` / `close` / `cancel` lifecycle over repeated request/reply exchanges rather than acting as a generic abstraction over all socket patterns. The session family is implemented and usable, but it should still be treated as experimental until session ownership is bound to a concrete multi-client identity model.

## Functions

```{r function-catalog, echo=FALSE, results='asis'}
knitr::asis_output(paste(readLines("function_catalog/functions.md"), collapse = "\n"))
```

## Build

```sh
make configure
make release
```

See `docs/protocol.md`, `docs/manifest.md`, `docs/security.md`, `docs/registry.md`, `docs/transports.md`, `docs/http.md`, and `docs/types.md` for the authoritative protocol, transport, TLS, HTTP-adapter, and session/query-family contract. `NEWS.md` summarizes notable landed changes, while `docs/design_review_checklist.md` is a better guide to the main remaining gaps and next implementation slices. TLS already supports both file-backed and in-memory PEM material through helpers such as `ducknng_tls_config_from_files(...)`, `ducknng_tls_config_from_pem(...)`, and `ducknng_self_signed_tls_config(...)`.

## Examples

### Start an IPC listener and inspect the registry

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'sql0',                         -- service name
  'ipc:///tmp/ducknng_sql0.ipc', -- listen URL
  1,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

SELECT name, listen, contexts, running, sessions
FROM ducknng_list_servers();

SELECT ducknng_stop_server('sql0');
```

### Request multiple REP contexts on one REP socket

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'sql_multi',                    -- service name
  'ipc:///tmp/ducknng_sql_multi.ipc', -- listen URL
  3,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

SELECT name, contexts, running
FROM ducknng_list_servers()
WHERE name = 'sql_multi';

SELECT ducknng_stop_server('sql_multi');
```

### DuckDB can also act as a client

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Start a local ducknng service that the following client examples will talk to.
SELECT ducknng_start_server(
  'sql_client_demo',              -- service name
  'ipc:///tmp/ducknng_sql_client_demo.ipc', -- listen URL
  1,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

-- The default RPC surface is minimal: manifest is built in, exec is opt-in.
SELECT * FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT);
SELECT * FROM ducknng_list_methods();

-- Register exec explicitly before exposing SQL execution over RPC.
SELECT ducknng_register_exec_method();
SELECT * FROM ducknng_list_methods();
SELECT * FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT);

-- RPC helper: run non-row statements and keep errors in-band.
SELECT * FROM ducknng_run_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'CREATE TABLE IF NOT EXISTS client_side_demo(i INTEGER)', 0::UBIGINT);
SELECT * FROM ducknng_run_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'INSERT INTO client_side_demo VALUES (10), (11)', 0::UBIGINT);

-- RPC helper: fetch row results through the unary query path.
SELECT * FROM ducknng_query_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'SELECT i, i > 10 AS gt_10 FROM client_side_demo ORDER BY i', 0::UBIGINT);

-- Primitive transport layer: open a req socket handle, dial it, and inspect the registry.
SELECT ducknng_open_socket('req');
SELECT ducknng_dial_socket(1, 'ipc:///tmp/ducknng_sql_client_demo.ipc', 1000);
SELECT * FROM ducknng_list_sockets();

-- Primitive transport layer: send the built-in manifest request frame.
-- The hex literal here is the current wire-format request for the always-on manifest method.
SELECT *
FROM ducknng_request_socket(
  1::UBIGINT,                                        -- socket_id
  from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
  1000                                               -- timeout_ms
);

SELECT *
FROM ducknng_request(
  'ipc:///tmp/ducknng_sql_client_demo.ipc',          -- url
  from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
  1000,                                              -- timeout_ms
  0::UBIGINT                                         -- tls_config_id
);

-- Raw scalar forms now mean raw reply frames as BLOBs.
SELECT substr(
  hex(
    ducknng_request_socket_raw(
      1,                                             -- socket_id
      from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
      1000                                           -- timeout_ms
    )
  ),
  1,
  28
);

-- Decode raw manifest and exec replies into envelope fields plus extracted text payload.
SELECT ok, version, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(ducknng_get_rpc_manifest_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT));

SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_run_rpc_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', 'CREATE TABLE IF NOT EXISTS client_side_demo(i INTEGER)', 0::UBIGINT)
);

-- The generic raw request helper can be decoded the same way.
SELECT ok, version, type_name, name, payload_text
FROM ducknng_decode_frame(
  ducknng_request_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', from_hex('01000000000000000000000000000000000000000000'), 1000, 0::UBIGINT)
);

-- Close the client socket handle and stop the demo server.
SELECT ducknng_close_socket(1);
SELECT ducknng_stop_server('sql_client_demo');
```

### Use `ducknng_ncurl()` against a local `nanonext` HTTPS server

A tiny local R `nanonext` HTTPS server is started during README rendering so this example runs without depending on the public internet. The core server definition is shown below because the transport helper and its carrier behavior are part of the example rather than hidden setup. The hidden setup only launches the script in the background and manages the temporary PID, log, and CA files needed while rendering. `ducknng_ncurl(...)` is the low-level HTTP/HTTPS transport primitive; the higher-level request, RPC, and session helpers have not been auto-routed over `http://` and `https://` yet.

```{r http-client-demo-setup, echo=FALSE, results='asis'}
pidfile <- "/tmp/ducknng_readme_http_demo.pid"
script <- "/tmp/ducknng_readme_http_demo.R"
logfile <- "/tmp/ducknng_readme_http_demo.log"
cafile <- "/tmp/ducknng_readme_http_demo_ca.pem"
if (file.exists(pidfile)) {
  pid <- suppressWarnings(as.integer(readLines(pidfile, warn = FALSE)))
  if (!is.na(pid)) tools::pskill(pid)
  unlink(pidfile)
}
http_demo_code <- paste(
  "library(nanonext)",
  "",
  "`%||%` <- function(x, y) if (is.null(x) || !length(x) || !nzchar(x)) y else x",
  "",
  "cert <- write_cert(cn = '127.0.0.1')",
  sprintf("writeLines(cert$client[[1]], %s)", deparse(cafile)),
  "",
  "server <- http_server(",
  "  url = 'https://127.0.0.1:18443',",
  "  handlers = list(",
  "    handler('/hello', function(req) {",
  "      list(status = 200L, headers = c('Content-Type' = 'text/plain', 'X-Test' = 'hello'), body = 'hello from ducknng https')",
  "    }),",
  "    handler('/echo', function(req) {",
  "      list(",
  "        status = 200L,",
  "        headers = c('Content-Type' = (req$headers[['Content-Type']] %||% 'application/octet-stream'), 'X-Test' = 'echo'),",
  "        body = req$body",
  "      )",
  "    }, method = 'POST')",
  "  ),",
  "  tls = tls_config(server = cert$server)",
  ")",
  sep = "\n"
)
cat("```r\n", http_demo_code, "\n```\n\n", sep = "")
launch_code <- paste(
  http_demo_code,
  "",
  "library(later)",
  "attr(server, 'start')()",
  "repeat later::run_now(Inf)",
  sep = "\n"
)
writeLines(launch_code, script)
status <- system(sprintf(
  "Rscript %s > %s 2>&1 & echo $! > %s",
  shQuote(script),
  shQuote(logfile),
  shQuote(pidfile)
))
if (status != 0 || !file.exists(pidfile)) {
  stop("failed to start the README HTTP demo server")
}
Sys.sleep(1)
pid <- suppressWarnings(as.integer(readLines(pidfile, warn = FALSE)))
if (is.na(pid) || system(sprintf("kill -0 %d >/dev/null 2>&1", pid)) != 0) {
  log_text <- if (file.exists(logfile)) paste(readLines(logfile, warn = FALSE), collapse = "\n") else ""
  stop("README HTTP demo server did not stay running\n", log_text)
}
```

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Register a client TLS handle that trusts the self-signed HTTPS demo server.
SELECT ducknng_tls_config_from_files(
  NULL,                                     -- cert_key_file
  '/tmp/ducknng_readme_http_demo_ca.pem',   -- ca_file
  NULL,                                     -- password
  2                                         -- auth_mode = require certificate validation
);

-- GET defaults are allowed: NULL method means GET and NULL body means no request body.
SELECT ok, status, body_text
FROM ducknng_ncurl(
  'https://127.0.0.1:18443/hello', -- url
  NULL,                            -- method
  NULL,                            -- headers_json
  NULL,                            -- body
  2000,                            -- timeout_ms
  1::UBIGINT                       -- tls_config_id
);

-- POST can send raw bytes while still exposing HTTP headers and status in-band.
SELECT ok, status, hex(body) AS body_hex, position('X-Test' IN headers_json) > 0 AS has_header
FROM ducknng_ncurl(
  'https://127.0.0.1:18443/echo',                        -- url
  'POST',                                                -- method
  '[{"name":"Content-Type","value":"application/octet-stream"}]', -- headers_json
  from_hex('01020304'),                                  -- body
  2000,                                                  -- timeout_ms
  1::UBIGINT                                             -- tls_config_id
);

-- Drop the temporary client TLS handle after the HTTPS checks complete.
SELECT ducknng_drop_tls_config(1);
```

```{r http-client-demo-cleanup, include=FALSE}
pidfile <- "/tmp/ducknng_readme_http_demo.pid"
if (file.exists(pidfile)) {
  pid <- suppressWarnings(as.integer(readLines(pidfile, warn = FALSE)))
  if (!is.na(pid)) tools::pskill(pid)
  unlink(pidfile)
}
unlink(c(
  "/tmp/ducknng_readme_http_demo.R",
  "/tmp/ducknng_readme_http_demo.log",
  "/tmp/ducknng_readme_http_demo_ca.pem"
))
```

### Launch raw socket send/recv airos and inspect send status explicitly

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Open one listening pair socket and one dialed peer.
SELECT ducknng_open_socket('pair');
SELECT ducknng_listen_socket(1, 'ipc:///tmp/ducknng_sql_pair_aio_demo.ipc', 134217728, 0::UBIGINT);
SELECT ducknng_open_socket('pair');
SELECT ducknng_dial_socket(2, 'ipc:///tmp/ducknng_sql_pair_aio_demo.ipc', 1000);

-- Start one async receive and one async send.
CREATE TEMP TABLE pair_recv AS
SELECT ducknng_recv_socket_raw_aio(1::UBIGINT, 1000) AS recv_aio;

CREATE TEMP TABLE pair_send AS
SELECT ducknng_send_socket_raw_aio(2::UBIGINT, from_hex('6173796e632072657175657374'), 1000) AS send_aio;

-- Collect the terminal send result. Send-only airos succeed with frame = NULL.
SELECT aio_id, ok, frame IS NULL AS no_frame
FROM ducknng_aio_collect((SELECT list_value(send_aio) FROM pair_send), 1000);

-- Collect the received raw frame from the peer side.
SELECT aio_id, ok, hex(frame) = '6173796E632072657175657374' AS got_payload
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM pair_recv), 1000);

-- Inspect whether the send phase completed successfully.
SELECT kind, state, phase, send_done, send_ok, recv_done, recv_ok IS NULL AS recv_ok_is_null
FROM ducknng_aio_status((SELECT send_aio FROM pair_send));

-- Clean up the terminal aio handles, temp state, and socket handles.
SELECT ducknng_aio_drop((SELECT send_aio FROM pair_send));
SELECT ducknng_aio_drop((SELECT recv_aio FROM pair_recv));
DROP TABLE pair_send;
DROP TABLE pair_recv;
SELECT ducknng_close_socket(2);
SELECT ducknng_close_socket(1);
```

### Launch raw requests asynchronously and collect the reply frames later

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Start a local listener that the aio requests will target.
SELECT ducknng_start_server(
  'sql_aio_demo',                -- service name
  'ipc:///tmp/ducknng_sql_aio_demo.ipc', -- listen URL
  1,                             -- REP contexts
  134217728,                     -- recv_max_bytes
  300000,                        -- session_idle_ms
  0                              -- tls_config_id (0 means plaintext)
);

-- Launch two raw request/reply futures and keep their aio ids in one temp row.
-- The hex literal below is the built-in manifest request frame in the current wire format.
CREATE TEMP TABLE aio_demo AS
SELECT
  ducknng_request_raw_aio(
    'ipc:///tmp/ducknng_sql_aio_demo.ipc', -- url
    from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
    1000,                                  -- timeout_ms
    0::UBIGINT                             -- tls_config_id
  ) AS aio1,
  ducknng_request_raw_aio(
    'ipc:///tmp/ducknng_sql_aio_demo.ipc', -- url
    from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
    1000,                                  -- timeout_ms
    0::UBIGINT                             -- tls_config_id
  ) AS aio2;

-- Confirm that both aio launches returned runtime handle ids.
SELECT aio1 > 0 AS aio1_started, aio2 > aio1 AS aio2_started_after_aio1
FROM aio_demo;

-- Wait for terminal results and return the full reply frames for later decoding.
SELECT aio_id, ok, octet_length(frame) > 0 AS has_frame
FROM ducknng_aio_collect((SELECT list_value(aio1, aio2) FROM aio_demo), 1000)
ORDER BY aio_id;

-- A collected aio is terminal, so ready() stays true until the handle is dropped.
SELECT ducknng_aio_ready(aio1) AS aio1_ready, ducknng_aio_ready(aio2) AS aio2_ready
FROM aio_demo;

-- Drop the terminal aio handles explicitly because SQL has no destructor hook.
SELECT ducknng_aio_drop(aio1) AND ducknng_aio_drop(aio2) AS dropped
FROM aio_demo;

-- Remove the temp state and stop the demo listener.
DROP TABLE aio_demo;
SELECT ducknng_stop_server('sql_aio_demo');
```

### Open, fetch, and close a query session explicitly

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Start a listener that will own the server-side query session.
SELECT ducknng_start_server(
  'sql_session_demo',            -- service name
  'ipc:///tmp/ducknng_sql_session_demo.ipc', -- listen URL
  1,                             -- REP contexts
  134217728,                     -- recv_max_bytes
  300000,                        -- session_idle_ms
  0                              -- tls_config_id (0 means plaintext)
);

-- Open one server-owned query session.
-- batch_rows = 0 and batch_bytes = 0 mean "use the server defaults" for this demo.
SELECT *
FROM ducknng_open_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc', -- url
  'SELECT 1 AS id UNION ALL SELECT 2 AS id ORDER BY id', -- SQL text to run remotely
  0::UBIGINT,                                -- batch_rows
  0::UBIGINT,                                -- batch_bytes
  0::UBIGINT                                 -- tls_config_id
);

-- The first fetch returns one Arrow IPC batch in payload.
SELECT ok, session_id, state IS NULL AS state_is_null, octet_length(payload) > 0 AS has_payload, end_of_stream
FROM ducknng_fetch_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc', -- url
  1::UBIGINT,                                -- session_id
  0::UBIGINT,                                -- batch_rows override
  0::UBIGINT,                                -- batch_bytes override
  0::UBIGINT                                 -- tls_config_id
);

-- The second fetch returns JSON control metadata saying the session is exhausted.
SELECT ok, session_id, state, payload IS NULL AS no_payload, end_of_stream
FROM ducknng_fetch_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc', -- url
  1::UBIGINT,                                -- session_id
  0::UBIGINT,                                -- batch_rows override
  0::UBIGINT,                                -- batch_bytes override
  0::UBIGINT                                 -- tls_config_id
);

-- Close the exhausted session explicitly.
SELECT *
FROM ducknng_close_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc', -- url
  1::UBIGINT,                                -- session_id
  0::UBIGINT                                 -- tls_config_id
);

-- Stop the demo listener after the session is closed.
SELECT ducknng_stop_server('sql_session_demo');
```

### `tls+tcp://` with a self-signed development TLS config

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Generate a self-signed loopback certificate and keep it as a runtime TLS handle.
SELECT ducknng_self_signed_tls_config('127.0.0.1', 365, 0);

-- Start a TLS listener using the generated config handle.
SELECT ducknng_start_server(
  'tls_demo_self',               -- service name
  'tls+tcp://127.0.0.1:45453',   -- listen URL
  1,                             -- REP contexts
  134217728,                     -- recv_max_bytes
  300000,                        -- session_idle_ms
  1                              -- tls_config_id returned above
);

-- Send the built-in manifest request frame over TLS and decode the reply.
SELECT ok, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'tls+tcp://127.0.0.1:45453',                -- url
    from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
    1000,                                       -- timeout_ms
    1                                           -- tls_config_id
  )
);

-- Clean up the TLS demo server and config handle.
SELECT ducknng_stop_server('tls_demo_self');
SELECT ducknng_drop_tls_config(1);
```

### `tls+tcp://` from file-backed certificate material

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Register a file-backed TLS config using committed loopback test certificates.
SELECT ducknng_tls_config_from_files(
  'test/certs/loopback-cert-key.pem', -- cert_chain_path
  'test/certs/loopback-ca.pem',       -- ca_cert_path
  NULL,                               -- private_key_path (included in the cert/key PEM above)
  0                                   -- mode flags
);

-- Start a TLS listener with that file-backed config.
SELECT ducknng_start_server(
  'tls_demo_files',              -- service name
  'tls+tcp://127.0.0.1:45454',   -- listen URL
  1,                             -- REP contexts
  134217728,                     -- recv_max_bytes
  300000,                        -- session_idle_ms
  1                              -- tls_config_id returned above
);

-- Inspect the manifest reply over TLS using the same config handle on the client side.
SELECT ok, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'tls+tcp://127.0.0.1:45454',                -- url
    from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
    1000,                                       -- timeout_ms
    1                                           -- tls_config_id
  )
);

-- Clean up the file-backed TLS demo server and config handle.
SELECT ducknng_stop_server('tls_demo_files');
SELECT ducknng_drop_tls_config(1);
```

### REQ/REP `EXEC` via `nanonext` as an interop example

This example is easier to follow as a short sequence: define the frame helpers, define the retry helpers, start a local server, run a few remote `exec` requests, inspect the same table locally, and then clean up.

#### Frame encoding and decoding helpers

```{r exec-example-frame-helpers}
# Little-endian helpers for the versioned ducknng RPC envelope.
write_u32le <- function(x) {
  writeBin(as.integer(x), raw(), size = 4L, endian = "little")
}

write_u64le <- function(x) {
  x <- as.double(x)
  c(write_u32le(x %% 2^32), write_u32le(floor(x / 2^32)))
}

read_u32le_bytes <- function(x) {
  sum(as.double(as.integer(x)) * 256^(0:3))
}

read_u64le_bytes <- function(x) {
  read_u32le_bytes(x[1:4]) + 2^32 * read_u32le_bytes(x[5:8])
}

read_u32le <- function(buf, offset) {
  read_u32le_bytes(buf[offset + 0:3])
}

read_u64le <- function(buf, offset) {
  read_u64le_bytes(buf[offset + 0:7])
}

# Encode the Arrow IPC payload carried by the built-in exec RPC method.
encode_ducknng_exec_payload <- function(sql, want_result = FALSE) {
  raw_con <- rawConnection(raw(), open = "r+")
  on.exit(close(raw_con), add = TRUE)

  nanoarrow::write_nanoarrow(
    data.frame(sql = sql, want_result = want_result),
    raw_con
  )

  rawConnectionValue(raw_con)
}

# Build a version 1 RPC call frame for the named method.
encode_ducknng_call_frame <- function(method, payload, flags = 0L) {
  method_name <- charToRaw(method)

  c(
    as.raw(1L),
    as.raw(1L),
    write_u32le(flags),
    write_u32le(length(method_name)),
    write_u32le(0),
    write_u64le(length(payload)),
    method_name,
    payload
  )
}

encode_ducknng_exec_request <- function(sql, want_result = FALSE) {
  encode_ducknng_call_frame(
    method = "exec",
    payload = encode_ducknng_exec_payload(sql, want_result = want_result)
  )
}

# Decode a reply envelope and parse its Arrow IPC payload when present.
decode_ducknng_exec_reply <- function(buf) {
  header_size <- 23L

  if (!is.raw(buf) || length(buf) < (header_size - 1L)) {
    stop("ducknng reply frame was empty or truncated", call. = FALSE)
  }

  name_len <- read_u32le(buf, 7)
  error_len <- read_u32le(buf, 11)
  payload_len <- read_u64le(buf, 15)

  name_start <- header_size
  error_start <- name_start + name_len
  payload_start <- error_start + error_len
  payload_end <- payload_start + payload_len - 1L

  payload <- if (payload_len > 0) buf[payload_start:payload_end] else raw()

  list(
    version = as.integer(buf[1]),
    type = as.integer(buf[2]),
    flags = read_u32le(buf, 3),
    method = rawToChar(buf[name_start:(error_start - 1L)]),
    error = if (error_len > 0) rawToChar(buf[error_start:(payload_start - 1L)]) else "",
    payload = payload,
    data = if (payload_len > 0) as.data.frame(nanoarrow::read_nanoarrow(payload)) else NULL
  )
}
```

#### REQ/REP retry helpers

```{r exec-example-retry-helpers}
# Send and receive with the same retry style used in mangoro examples.
send_ducknng_frame <- function(socket, frame, timeout_ms = 1000L, max_attempts = 20L) {
  attempt <- 1L
  send_result <- nanonext::send(socket, frame, mode = "raw", block = timeout_ms)

  while (nanonext::is_error_value(send_result) && attempt < max_attempts) {
    Sys.sleep(0.25)
    send_result <- nanonext::send(socket, frame, mode = "raw", block = timeout_ms)
    attempt <- attempt + 1L
  }

  send_result
}

recv_ducknng_frame <- function(socket, timeout_ms = 1000L, max_attempts = 20L) {
  attempt <- 1L
  response <- nanonext::recv(socket, mode = "raw", block = timeout_ms)

  while (
    (
      nanonext::is_error_value(response) ||
      !is.raw(response) ||
      length(response) < 22L
    ) &&
    attempt < max_attempts
  ) {
    Sys.sleep(0.25)
    response <- nanonext::recv(socket, mode = "raw", block = timeout_ms)
    attempt <- attempt + 1L
  }

  response
}

run_ducknng_req <- function(socket, frame, timeout_ms = 1000L) {
  send_ducknng_frame(socket, frame, timeout_ms = timeout_ms)
  decode_ducknng_exec_reply(
    recv_ducknng_frame(socket, timeout_ms = timeout_ms)
  )
}

# Wait for an IPC listener path to appear before dialing it.
wait_for_ducknng_listener <- function(path, timeout_secs = 10, interval_secs = 0.1) {
  deadline <- Sys.time() + timeout_secs

  while (Sys.time() < deadline) {
    if (file.exists(path)) {
      return(invisible(TRUE))
    }
    Sys.sleep(interval_secs)
  }

  stop("ducknng IPC listener did not become ready in time", call. = FALSE)
}
```

#### Start a local ducknng server and dial a req socket

```{r exec-example-setup}
# Start a ducknng server in this R session, then talk to it over a req socket.
ext_path <- normalizePath("build/release/ducknng.duckdb_extension")
ipc_path <- tempfile(pattern = "ducknng_readme_exec_", tmpdir = "/tmp", fileext = ".ipc")
ipc_url <- paste0("ipc://", ipc_path)

db_driver <- duckdb::duckdb(
  dbdir = ":memory:",
  config = list(allow_unsigned_extensions = "true")
)
db_con <- DBI::dbConnect(db_driver)

DBI::dbGetQuery(db_con, "SELECT 42 AS ok")
DBI::dbExecute(db_con, sprintf("LOAD '%s'", ext_path))
DBI::dbGetQuery(
  db_con,
  sprintf(
    "SELECT ducknng_start_server('sql_exec', '%s', 1, 134217728, 300000, 0::UBIGINT)",
    ipc_url
  )
)
DBI::dbGetQuery(db_con, "SELECT ducknng_register_exec_method()")

wait_for_ducknng_listener(ipc_path)
req <- nanonext::socket("req", dial = ipc_url)
```

#### Create and insert rows remotely

```{r exec-example-mutate}
# Create a table remotely and inspect the metadata reply.
create_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request("CREATE TABLE ducknng_exec_demo(i INTEGER)")
)
create_reply$data

# Insert rows remotely and inspect the second metadata reply.
insert_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request("INSERT INTO ducknng_exec_demo VALUES (42), (99)")
)
insert_reply$data
```

#### Query rows remotely with `want_result = TRUE`

```{r exec-example-query}
select_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request(
    "SELECT i, i > 50 AS gt_50 FROM ducknng_exec_demo ORDER BY i",
    want_result = TRUE
  )
)
select_reply$data
```

#### Inspect the same table locally through DuckDB

```{r exec-example-local}
server_rows <- DBI::dbGetQuery(db_con, "SELECT * FROM ducknng_exec_demo ORDER BY i")
server_rows
```

#### Stop the server and clean up

```{r exec-example-cleanup}
DBI::dbGetQuery(db_con, "SELECT ducknng_stop_server('sql_exec')")
close(req)
DBI::dbDisconnect(db_con)
unlink(ipc_path)
```

## References

- [NNG](https://nng.nanomsg.org/) for the underlying messaging library and REQ/REP transport model.
- [`r-lib/nanonext`](https://github.com/r-lib/nanonext) for the main client/server ergonomics reference.
- [`sounkou-bioinfo/mangoro`](https://github.com/sounkou-bioinfo/mangoro) for the thin-envelope + Arrow IPC RPC direction.
- [DuckDB C API](https://duckdb.org/docs/stable/clients/c/api) for the extension and SQL integration boundary.
- [Apache Arrow IPC](https://arrow.apache.org/docs/format/Columnar.html#serialization-and-interprocess-communication-ipc) for the tabular payload format.

