---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::knit_engines$set(sql = function(options) {
  code <- paste(options$code, collapse = "\n")
  extension_path <- normalizePath("build/release/ducknng.duckdb_extension", mustWork = FALSE)
  extension_path <- gsub("'", "''", extension_path, fixed = TRUE)
  code <- paste(sprintf("LOAD '%s';", extension_path), code, sep = "\n")
  if (isTRUE(options$eval)) {
    out <- system2("duckdb", "-unsigned", input = code, stdout = TRUE, stderr = TRUE)
    status <- attr(out, "status")
    if (!is.null(status) && status != 0) stop(paste(out, collapse = "\n"))
    knitr::engine_output(options, code, out)
  } else {
    knitr::engine_output(options, code, NULL)
  }
})
```

# ducknng

`ducknng` is a pure C DuckDB extension for DuckDB + NNG REQ/REP interop.

Today it provides:

- named server start/stop from SQL
- registry introspection with `ducknng_servers()`
- one REP socket with one or more REP contexts
- a working `EXEC` request/reply path over the raw wire protocol

## Functions

```{r function-catalog, echo=FALSE, results='asis'}
knitr::asis_output(paste(readLines("function_catalog/functions.md"), collapse = "\n"))
```

## Build

```sh
make configure
make release
```

See also `NEWS.md` for the current implementation status and planned next steps.

## Examples

### Start an IPC listener and inspect the registry

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_server_start(
  'sql0',
  'ipc:///tmp/ducknng_sql0.ipc',
  1,
  134217728,
  300000,
  NULL,
  NULL,
  NULL
);

SELECT name, listen, contexts, running, sessions
FROM ducknng_servers();

SELECT ducknng_server_stop('sql0');
```

### Request multiple REP contexts on one REP socket

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_server_start(
  'sql_multi',
  'ipc:///tmp/ducknng_sql_multi.ipc',
  3,
  134217728,
  300000,
  NULL,
  NULL,
  NULL
);

SELECT name, contexts, running
FROM ducknng_servers()
WHERE name = 'sql_multi';

SELECT ducknng_server_stop('sql_multi');
```

### REQ/REP `EXEC` via `nanonext` as an interop example

```{r exec-example}
suppressWarnings(suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
  library(nanonext)
  library(nanoarrow)
}))

# Little-endian integer encoders for the versioned ducknng RPC envelope.
u32le <- function(x) writeBin(as.integer(x), raw(), size = 4L, endian = "little")
u64le <- function(x) {
  x <- as.double(x)
  c(u32le(x %% 2^32), u32le(floor(x / 2^32)))
}
le_u32_from_raw <- function(x) sum(as.double(as.integer(x)) * 256^(0:3))
le_u64_from_raw <- function(x) le_u32_from_raw(x[1:4]) + 2^32 * le_u32_from_raw(x[5:8])
read_u32le <- function(buf, offset) le_u32_from_raw(buf[offset + 0:3])
read_u64le <- function(buf, offset) le_u64_from_raw(buf[offset + 0:7])

# Serialize a one-row exec request as Arrow IPC using nanoarrow.
encode_ducknng_exec_request <- function(sql, want_result = FALSE) {
  con <- rawConnection(raw(), open = "r+")
  on.exit(close(con))
  nanoarrow::write_nanoarrow(
    data.frame(sql = sql, want_result = want_result),
    con
  )
  payload <- rawConnectionValue(con)
  name <- charToRaw("exec")
  c(
    as.raw(1L),                  # envelope version
    as.raw(1L),                  # message type: call
    u32le(0),                    # flags
    u32le(length(name)),         # method-name length
    u32le(0),                    # error-string length on requests
    u64le(length(payload)),      # payload length
    name,
    payload
  )
}

# Decode a reply envelope and, when present, parse the Arrow IPC payload.
decode_ducknng_exec_reply <- function(buf) {
  name_len <- read_u32le(buf, 7)
  error_len <- read_u32le(buf, 11)
  payload_len <- read_u64le(buf, 15)
  name_start <- 23L
  error_start <- name_start + name_len
  payload_start <- error_start + error_len
  payload_end <- payload_start + payload_len - 1L
  payload <- if (payload_len > 0) buf[payload_start:payload_end] else raw()
  list(
    version = as.integer(buf[1]),
    type = as.integer(buf[2]),
    flags = read_u32le(buf, 3),
    method = rawToChar(buf[name_start:(error_start - 1L)]),
    error = if (error_len > 0) rawToChar(buf[error_start:(payload_start - 1L)]) else "",
    payload = payload,
    data = if (payload_len > 0) as.data.frame(nanoarrow::read_nanoarrow(payload)) else NULL
  )
}

# Use a temporary IPC path and launch a DuckDB-backed ducknng server in a child process.
ext_path <- normalizePath("build/release/ducknng.duckdb_extension")
ipc_path <- tempfile(pattern = "ducknng_readme_exec_", tmpdir = "/tmp", fileext = ".ipc")
ipc_url <- paste0("ipc://", ipc_path)

server_job <- parallel::mcparallel({
  drv <- duckdb::duckdb(config = list(allow_unsigned_extensions = "true"))
  con <- DBI::dbConnect(drv, dbdir = ":memory:")
  DBI::dbExecute(con, sprintf("LOAD '%s'", ext_path))
  DBI::dbGetQuery(con, sprintf(
    "SELECT ducknng_server_start('sql_exec', '%s', 1, 134217728, 300000, NULL, NULL, NULL)",
    ipc_url
  ))
  Sys.sleep(2)
  rows <- DBI::dbGetQuery(con, "SELECT * FROM ducknng_exec_demo ORDER BY i")
  DBI::dbGetQuery(con, "SELECT ducknng_server_stop('sql_exec')")
  DBI::dbDisconnect(con, shutdown = TRUE)
  rows
})

# Give the listener a moment to come up, then dial it with a normal req socket.
Sys.sleep(1)
req <- nanonext::socket("req", dial = ipc_url, autostart = NA)

# Create a table remotely and inspect the reply metadata.
nanonext::send(req, encode_ducknng_exec_request("CREATE TABLE ducknng_exec_demo(i INTEGER)"), mode = "raw", block = 1000L)
create_reply <- decode_ducknng_exec_reply(nanonext::recv(req, mode = "raw", block = 1000L))
create_reply$data

# Insert rows remotely and inspect the second reply.
nanonext::send(req, encode_ducknng_exec_request("INSERT INTO ducknng_exec_demo VALUES (42), (99)"), mode = "raw", block = 1000L)
insert_reply <- decode_ducknng_exec_reply(nanonext::recv(req, mode = "raw", block = 1000L))
insert_reply$data

# Collect the server-side query result from the child process.
server_rows <- parallel::mccollect(server_job)[[1]]
server_rows

close(req)
```

