---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::knit_engines$set(sql = function(options) {
  code <- paste(options$code, collapse = "\n")
  extension_path <- normalizePath("build/release/ducknng.duckdb_extension", mustWork = FALSE)
  extension_path <- gsub("'", "''", extension_path, fixed = TRUE)
  code <- paste(sprintf("LOAD '%s';", extension_path), code, sep = "\n")
  if (isTRUE(options$eval)) {
    out <- system2("duckdb", "-unsigned", input = code, stdout = TRUE, stderr = TRUE)
    status <- attr(out, "status")
    if (!is.null(status) && status != 0) stop(paste(out, collapse = "\n"))
    knitr::engine_output(options, code, out)
  } else {
    knitr::engine_output(options, code, NULL)
  }
})
```

# ducknng

`ducknng` is a pure C DuckDB extension for DuckDB + NNG REQ/REP interop.

Its SQL-facing client and server interface is explicitly modeled over [`r-lib/nanonext`](https://github.com/r-lib/nanonext), taking req/rep sockets, dial/listen semantics, context-oriented NNG usage, and a practical messaging surface as the reference point for what should feel natural from DuckDB SQL. Its RPC framing and tabular payload direction are likewise informed by Arrow IPC based RPC work such as [`sounkou-bioinfo/mangoro`](https://github.com/sounkou-bioinfo/mangoro) and related projects, where a thin envelope is kept separate from Arrow IPC request and reply payloads instead of being buried inside one-off method-specific binaries.

At the moment, `ducknng` already exposes a low-level transport layer with socket-open, dial, close, and raw request primitives, alongside SQL-visible server lifecycle control and runtime introspection through `ducknng_start_server()`, `ducknng_stop_server()`, and `ducknng_list_servers()`. That transport surface is operation-oriented rather than transport-specific: transport is autodetected from the URL scheme in the same general style as `nanonext`, so the same API can work with `inproc://`, `ipc://`, `tcp://`, and `tls+tcp://` without spawning a separate helper family for each scheme, while TLS material is passed through reusable `tls_config_id` handles instead of inline transport-specific arguments. The current SQL client socket-handle layer is still honest but narrow: `ducknng_open_socket()` only implements `req` today. On the server side it runs one REP socket with one or more REP contexts, and on top of that transport it already supports a working RPC request/reply path with manifest discovery, metadata-oriented execution, and unary row-returning execution over the current raw wire and Arrow IPC model.

A SQL-visible aio surface now exists for raw request/reply futures. In `nanonext` terms, an aio here means one pending NNG operation backed by `nng_aio`, not a new wire mode or a background job queue. The first exposed slice is deliberately raw-first: `ducknng_request_raw_aio()` and `ducknng_request_socket_raw_aio()` launch framed requests and return aio handle ids immediately, `ducknng_aio_ready()` reports whether a handle has reached a terminal state, `ducknng_aio_collect()` waits for and returns full reply frames, and `ducknng_aio_drop()` releases terminal handles explicitly.

The session query family is now exposed on the SQL client side through `ducknng_open_query()`, `ducknng_fetch_query()`, `ducknng_close_query()`, and `ducknng_cancel_query()`. That surface stays control-first: opening, closing, and cancelling a session return JSON-derived control rows, while `ducknng_fetch_query()` returns either the next Arrow IPC batch as a BLOB payload or an end-of-stream control row. Until a real owner-identity model is implemented, that family should still be treated as development scaffolding rather than as a production-safe multi-client surface.

The transport-security utility layer now has a runtime model for TLS configuration handles. Those handles can be assembled from file-backed certificate material, from in-memory PEM strings, or from self-signed development certificates generated on demand. The actual `tls+tcp://` transport wiring remains isolated behind `ducknng_nng_compat.c`, but it is now enabled for real loopback listener and one-shot client-request paths so the public API no longer needs a separate transport-specific helper family. This follows the same broad direction as `nanonext`, which exposes TLS helper facilities backed by mbedTLS rather than forcing every client to hand-roll certificate setup outside the transport layer.

## Functions

```{r function-catalog, echo=FALSE, results='asis'}
knitr::asis_output(paste(readLines("function_catalog/functions.md"), collapse = "\n"))
```

## Build

```sh
make configure
make release
```

See also `NEWS.md` for the current implementation status and planned next steps, and `docs/protocol.md`, `docs/manifest.md`, `docs/security.md`, `docs/registry.md`, and `docs/types.md` for the binding transport, TLS, and session/query-family contract. Those design docs also define the current TLS direction: certificate and key material can be accepted either from files or from in-memory PEM content, with helper utilities for development certificate generation and client/server TLS configuration rather than a file-only setup path.

## Examples

### Start an IPC listener and inspect the registry

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'sql0',                         -- service name
  'ipc:///tmp/ducknng_sql0.ipc', -- listen URL
  1,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

SELECT name, listen, contexts, running, sessions
FROM ducknng_list_servers();

SELECT ducknng_stop_server('sql0');
```

### Request multiple REP contexts on one REP socket

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'sql_multi',                    -- service name
  'ipc:///tmp/ducknng_sql_multi.ipc', -- listen URL
  3,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

SELECT name, contexts, running
FROM ducknng_list_servers()
WHERE name = 'sql_multi';

SELECT ducknng_stop_server('sql_multi');
```

### DuckDB can also act as a client

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Start a local ducknng service that the following client examples will talk to.
SELECT ducknng_start_server(
  'sql_client_demo',              -- service name
  'ipc:///tmp/ducknng_sql_client_demo.ipc', -- listen URL
  1,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

-- The default RPC surface is minimal: manifest is built in, exec is opt-in.
SELECT * FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT);
SELECT * FROM ducknng_list_methods();

-- Register exec explicitly before exposing SQL execution over RPC.
SELECT ducknng_register_exec_method();
SELECT * FROM ducknng_list_methods();
SELECT * FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT);

-- RPC helper: run non-row statements and keep errors in-band.
SELECT * FROM ducknng_run_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'CREATE TABLE IF NOT EXISTS client_side_demo(i INTEGER)', 0::UBIGINT);
SELECT * FROM ducknng_run_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'INSERT INTO client_side_demo VALUES (10), (11)', 0::UBIGINT);

-- RPC helper: fetch row results through the unary query path.
SELECT * FROM ducknng_query_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'SELECT i, i > 10 AS gt_10 FROM client_side_demo ORDER BY i', 0::UBIGINT);

-- Primitive transport layer: open a socket handle, dial it, and inspect the registry.
SELECT ducknng_open_socket('req');
SELECT ducknng_dial_socket(1, 'ipc:///tmp/ducknng_sql_client_demo.ipc', 1000);
SELECT * FROM ducknng_list_sockets();

-- Primitive transport layer: send a raw manifest frame and inspect the reply prefix.
SELECT * FROM ducknng_request_socket(1::UBIGINT, from_hex('01000000000000000000000000000000000000000000'), 1000);
SELECT * FROM ducknng_request('ipc:///tmp/ducknng_sql_client_demo.ipc', from_hex('01000000000000000000000000000000000000000000'), 1000, 0::UBIGINT);

-- Raw scalar forms now mean raw reply frames as BLOBs.
SELECT substr(hex(ducknng_request_socket_raw(1, from_hex('01000000000000000000000000000000000000000000'), 1000)), 1, 28);

-- Decode raw manifest and exec replies into envelope fields plus extracted text payload.
SELECT ok, version, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(ducknng_get_rpc_manifest_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT));

SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_run_rpc_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', 'CREATE TABLE IF NOT EXISTS client_side_demo(i INTEGER)', 0::UBIGINT)
);

-- The generic raw request helper can be decoded the same way.
SELECT ok, version, type_name, name, payload_text
FROM ducknng_decode_frame(
  ducknng_request_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', from_hex('01000000000000000000000000000000000000000000'), 1000, 0::UBIGINT)
);

-- Close the client socket handle and stop the demo server.
SELECT ducknng_close_socket(1);
SELECT ducknng_stop_server('sql_client_demo');
```

### Launch raw requests asynchronously and collect the reply frames later

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'sql_aio_demo',
  'ipc:///tmp/ducknng_sql_aio_demo.ipc',
  1,
  134217728,
  300000,
  0
);

CREATE TEMP TABLE aio_demo AS
SELECT
  ducknng_request_raw_aio(
    'ipc:///tmp/ducknng_sql_aio_demo.ipc',
    from_hex('01000000000000000000000000000000000000000000'),
    1000,
    0::UBIGINT
  ) AS aio1,
  ducknng_request_raw_aio(
    'ipc:///tmp/ducknng_sql_aio_demo.ipc',
    from_hex('01000000000000000000000000000000000000000000'),
    1000,
    0::UBIGINT
  ) AS aio2;

SELECT aio1 > 0 AS aio1_started, aio2 > aio1 AS aio2_started_after_aio1
FROM aio_demo;

SELECT aio_id, ok, octet_length(frame) > 0 AS has_frame
FROM ducknng_aio_collect((SELECT list_value(aio1, aio2) FROM aio_demo), 1000)
ORDER BY aio_id;

SELECT ducknng_aio_ready(aio1) AS aio1_ready, ducknng_aio_ready(aio2) AS aio2_ready
FROM aio_demo;

SELECT ducknng_aio_drop(aio1) AND ducknng_aio_drop(aio2) AS dropped
FROM aio_demo;

DROP TABLE aio_demo;
SELECT ducknng_stop_server('sql_aio_demo');
```

### Open, fetch, and close a query session explicitly

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'sql_session_demo',
  'ipc:///tmp/ducknng_sql_session_demo.ipc',
  1,
  134217728,
  300000,
  0
);

-- Open one server-owned query session and inspect the control reply.
SELECT *
FROM ducknng_open_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc',
  'SELECT 1 AS id UNION ALL SELECT 2 AS id ORDER BY id',
  0::UBIGINT,
  0::UBIGINT,
  0::UBIGINT
);

-- The first fetch returns one Arrow IPC batch in payload.
SELECT ok, session_id, state IS NULL AS state_is_null, octet_length(payload) > 0 AS has_payload, end_of_stream
FROM ducknng_fetch_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc',
  1::UBIGINT,
  0::UBIGINT,
  0::UBIGINT,
  0::UBIGINT
);

-- The second fetch returns JSON control metadata saying the session is exhausted.
SELECT ok, session_id, state, payload IS NULL AS no_payload, end_of_stream
FROM ducknng_fetch_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc',
  1::UBIGINT,
  0::UBIGINT,
  0::UBIGINT,
  0::UBIGINT
);

-- Close the exhausted session explicitly.
SELECT *
FROM ducknng_close_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc',
  1::UBIGINT,
  0::UBIGINT
);

SELECT ducknng_stop_server('sql_session_demo');
```

### `tls+tcp://` with a self-signed development TLS config

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Generate a self-signed loopback certificate and keep it as a runtime TLS handle.
SELECT ducknng_self_signed_tls_config('127.0.0.1', 365, 0);

-- Start a TLS listener using the generated config handle.
SELECT ducknng_start_server(
  'tls_demo_self',
  'tls+tcp://127.0.0.1:45453',
  1,
  134217728,
  300000,
  1
);

-- Send a raw manifest request over TLS and decode the reply frame.
SELECT ok, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'tls+tcp://127.0.0.1:45453',
    from_hex('01000000000000000000000000000000000000000000'),
    1000,
    1
  )
);

-- Clean up the TLS demo server and config handle.
SELECT ducknng_stop_server('tls_demo_self');
SELECT ducknng_drop_tls_config(1);
```

### `tls+tcp://` from file-backed certificate material

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Register a file-backed TLS config using committed loopback test certificates.
SELECT ducknng_tls_config_from_files(
  'test/certs/loopback-cert-key.pem',
  'test/certs/loopback-ca.pem',
  NULL,
  0
);

-- Start a TLS listener with that file-backed config.
SELECT ducknng_start_server(
  'tls_demo_files',
  'tls+tcp://127.0.0.1:45454',
  1,
  134217728,
  300000,
  1
);

-- Inspect the manifest reply over TLS using the same config handle on the client side.
SELECT ok, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'tls+tcp://127.0.0.1:45454',
    from_hex('01000000000000000000000000000000000000000000'),
    1000,
    1
  )
);

-- Clean up the file-backed TLS demo server and config handle.
SELECT ducknng_stop_server('tls_demo_files');
SELECT ducknng_drop_tls_config(1);
```

### REQ/REP `EXEC` via `nanonext` as an interop example

```{r exec-example}
# Little-endian helpers for the versioned ducknng RPC envelope.
write_u32le <- function(x) {
  writeBin(as.integer(x), raw(), size = 4L, endian = "little")
}

write_u64le <- function(x) {
  x <- as.double(x)
  c(write_u32le(x %% 2^32), write_u32le(floor(x / 2^32)))
}

read_u32le_bytes <- function(x) {
  sum(as.double(as.integer(x)) * 256^(0:3))
}

read_u64le_bytes <- function(x) {
  read_u32le_bytes(x[1:4]) + 2^32 * read_u32le_bytes(x[5:8])
}

read_u32le <- function(buf, offset) {
  read_u32le_bytes(buf[offset + 0:3])
}

read_u64le <- function(buf, offset) {
  read_u64le_bytes(buf[offset + 0:7])
}

# Encode the Arrow IPC payload carried by the built-in exec RPC method.
encode_ducknng_exec_payload <- function(sql, want_result = FALSE) {
  raw_con <- rawConnection(raw(), open = "r+")
  on.exit(close(raw_con), add = TRUE)

  nanoarrow::write_nanoarrow(
    data.frame(sql = sql, want_result = want_result),
    raw_con
  )

  rawConnectionValue(raw_con)
}

# Build a version 1 RPC call frame for the named method.
encode_ducknng_call_frame <- function(method, payload, flags = 0L) {
  method_name <- charToRaw(method)

  c(
    as.raw(1L),
    as.raw(1L),
    write_u32le(flags),
    write_u32le(length(method_name)),
    write_u32le(0),
    write_u64le(length(payload)),
    method_name,
    payload
  )
}

encode_ducknng_exec_request <- function(sql, want_result = FALSE) {
  encode_ducknng_call_frame(
    method = "exec",
    payload = encode_ducknng_exec_payload(sql, want_result = want_result)
  )
}

# Decode a reply envelope and parse its Arrow IPC payload when present.
decode_ducknng_exec_reply <- function(buf) {
  header_size <- 23L

  if (!is.raw(buf) || length(buf) < (header_size - 1L)) {
    stop("ducknng reply frame was empty or truncated", call. = FALSE)
  }

  name_len <- read_u32le(buf, 7)
  error_len <- read_u32le(buf, 11)
  payload_len <- read_u64le(buf, 15)

  name_start <- header_size
  error_start <- name_start + name_len
  payload_start <- error_start + error_len
  payload_end <- payload_start + payload_len - 1L

  payload <- if (payload_len > 0) buf[payload_start:payload_end] else raw()

  list(
    version = as.integer(buf[1]),
    type = as.integer(buf[2]),
    flags = read_u32le(buf, 3),
    method = rawToChar(buf[name_start:(error_start - 1L)]),
    error = if (error_len > 0) rawToChar(buf[error_start:(payload_start - 1L)]) else "",
    payload = payload,
    data = if (payload_len > 0) as.data.frame(nanoarrow::read_nanoarrow(payload)) else NULL
  )
}

# Send and receive with the same retry style used in mangoro examples.
send_ducknng_frame <- function(socket, frame, timeout_ms = 1000L, max_attempts = 20L) {
  attempt <- 1L
  send_result <- nanonext::send(socket, frame, mode = "raw", block = timeout_ms)

  while (nanonext::is_error_value(send_result) && attempt < max_attempts) {
    Sys.sleep(0.25)
    send_result <- nanonext::send(socket, frame, mode = "raw", block = timeout_ms)
    attempt <- attempt + 1L
  }

  send_result
}

recv_ducknng_frame <- function(socket, timeout_ms = 1000L, max_attempts = 20L) {
  attempt <- 1L
  response <- nanonext::recv(socket, mode = "raw", block = timeout_ms)

  while (
    (
      nanonext::is_error_value(response) ||
      !is.raw(response) ||
      length(response) < 22L
    ) &&
    attempt < max_attempts
  ) {
    Sys.sleep(0.25)
    response <- nanonext::recv(socket, mode = "raw", block = timeout_ms)
    attempt <- attempt + 1L
  }

  response
}

run_ducknng_req <- function(socket, frame, timeout_ms = 1000L) {
  send_ducknng_frame(socket, frame, timeout_ms = timeout_ms)
  decode_ducknng_exec_reply(
    recv_ducknng_frame(socket, timeout_ms = timeout_ms)
  )
}

# Wait for an IPC listener path to appear before dialing it.
wait_for_ducknng_listener <- function(path, timeout_secs = 10, interval_secs = 0.1) {
  deadline <- Sys.time() + timeout_secs

  while (Sys.time() < deadline) {
    if (file.exists(path)) {
      return(invisible(TRUE))
    }
    Sys.sleep(interval_secs)
  }

  stop("ducknng IPC listener did not become ready in time", call. = FALSE)
}

# Start a ducknng server in this R session, then talk to it over a req socket.
ext_path <- normalizePath("build/release/ducknng.duckdb_extension")
ipc_path <- tempfile(pattern = "ducknng_readme_exec_", tmpdir = "/tmp", fileext = ".ipc")
ipc_url <- paste0("ipc://", ipc_path)

db_driver <- duckdb::duckdb(
  dbdir = ":memory:",
  config = list(allow_unsigned_extensions = "true")
)
db_con <- DBI::dbConnect(db_driver)

DBI::dbGetQuery(db_con, "SELECT 42 AS ok")
DBI::dbExecute(db_con, sprintf("LOAD '%s'", ext_path))
DBI::dbGetQuery(
  db_con,
  sprintf(
    "SELECT ducknng_start_server('sql_exec', '%s', 1, 134217728, 300000, 0::UBIGINT)",
    ipc_url
  )
)
DBI::dbGetQuery(db_con, "SELECT ducknng_register_exec_method()")

wait_for_ducknng_listener(ipc_path)
req <- nanonext::socket("req", dial = ipc_url)

# Create a table remotely and inspect the metadata reply.
create_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request("CREATE TABLE ducknng_exec_demo(i INTEGER)")
)
create_reply$data

# Insert rows remotely and inspect the second metadata reply.
insert_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request("INSERT INTO ducknng_exec_demo VALUES (42), (99)")
)
insert_reply$data

# Request result rows directly with want_result = TRUE.
select_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request(
    "SELECT i, i > 50 AS gt_50 FROM ducknng_exec_demo ORDER BY i",
    want_result = TRUE
  )
)
select_reply$data

# Inspect the same table locally through the DuckDB connection that owns the server.
server_rows <- DBI::dbGetQuery(db_con, "SELECT * FROM ducknng_exec_demo ORDER BY i")
server_rows

# Stop the server and clean up the socket and temporary IPC path.
DBI::dbGetQuery(db_con, "SELECT ducknng_stop_server('sql_exec')")
close(req)
DBI::dbDisconnect(db_con)
unlink(ipc_path)
```

