---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::knit_engines$set(sql = function(options) {
  code <- paste(options$code, collapse = "\n")
  extension_path <- normalizePath("build/release/ducknng.duckdb_extension", mustWork = FALSE)
  extension_path <- gsub("'", "''", extension_path, fixed = TRUE)
  code <- paste(sprintf("LOAD '%s';", extension_path), code, sep = "\n")
  if (isTRUE(options$eval)) {
    out <- system2("duckdb", "-unsigned", input = code, stdout = TRUE, stderr = TRUE)
    status <- attr(out, "status")
    if (!is.null(status) && status != 0) stop(paste(out, collapse = "\n"))
    knitr::engine_output(options, code, out)
  } else {
    knitr::engine_output(options, code, NULL)
  }
})
```

# ducknng

`ducknng` is a pure C DuckDB extension for DuckDB + NNG REQ/REP interop.

Its SQL-facing client and server interface is explicitly modeled over [`r-lib/nanonext`](https://github.com/r-lib/nanonext), taking req/rep sockets, dial/listen semantics, context-oriented NNG usage, and a practical messaging surface as the reference point for what should feel natural from DuckDB SQL. Its RPC framing and tabular payload direction are likewise informed by Arrow IPC based RPC work such as [`sounkou-bioinfo/mangoro`](https://github.com/sounkou-bioinfo/mangoro) and related projects, where a thin envelope is kept separate from Arrow IPC request and reply payloads instead of being buried inside one-off method-specific binaries.

At the moment, `ducknng` already exposes a low-level transport layer with socket-open, dial, close, and raw request primitives, alongside SQL-visible server lifecycle control and runtime introspection through `ducknng_start_server()`, `ducknng_stop_server()`, and `ducknng_list_servers()`. That transport surface is operation-oriented rather than transport-specific: transport is autodetected from the URL scheme in the same general style as `nanonext`, so the same API can work with `inproc://`, `ipc://`, `tcp://`, and later `tls+tcp://` without spawning a separate helper family for each scheme. On the server side it runs one REP socket with one or more REP contexts, and on top of that transport it already supports a working RPC request/reply path with manifest discovery, metadata-oriented execution, and unary row-returning execution over the current raw wire and Arrow IPC model.

The next documented protocol slice is the session query family: `query_open`, `fetch`, `close`, and `cancel`. In the current contract, `query_open` opens one server-owned query session and returns JSON control metadata, `fetch` is the only method that may return streamed row batches, `close` is the normal explicit cleanup path, and `cancel` is best-effort rather than a guarantee of immediate interruption. Until a real owner-identity model is implemented, that family should still be treated as development scaffolding rather than as a production-safe multi-client surface.

## Functions

```{r function-catalog, echo=FALSE, results='asis'}
knitr::asis_output(paste(readLines("function_catalog/functions.md"), collapse = "\n"))
```

## Build

```sh
make configure
make release
```

See also `NEWS.md` for the current implementation status and planned next steps, and `docs/protocol.md`, `docs/manifest.md`, `docs/security.md`, `docs/registry.md`, and `docs/types.md` for the binding transport, TLS, and session/query-family contract. Those design docs also define the intended TLS direction: certificate and key material should be accepted either from files or from in-memory PEM content, with helper utilities for development certificate generation and client/server TLS configuration rather than a file-only setup path.

## Examples

### Start an IPC listener and inspect the registry

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'sql0',
  'ipc:///tmp/ducknng_sql0.ipc',
  1,
  134217728,
  300000,
  NULL,
  NULL,
  NULL
);

SELECT name, listen, contexts, running, sessions
FROM ducknng_list_servers();

SELECT ducknng_stop_server('sql0');
```

### Request multiple REP contexts on one REP socket

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'sql_multi',
  'ipc:///tmp/ducknng_sql_multi.ipc',
  3,
  134217728,
  300000,
  NULL,
  NULL,
  NULL
);

SELECT name, contexts, running
FROM ducknng_list_servers()
WHERE name = 'sql_multi';

SELECT ducknng_stop_server('sql_multi');
```

### DuckDB can also act as a client

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Start a local ducknng service that the following client examples will talk to.
SELECT ducknng_start_server(
  'sql_client_demo',
  'ipc:///tmp/ducknng_sql_client_demo.ipc',
  1,
  134217728,
  300000,
  NULL,
  NULL,
  NULL
);

-- RPC helper: fetch the manifest as a structured status row.
SELECT * FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_sql_client_demo.ipc');

-- RPC helper: run non-row statements and keep errors in-band.
SELECT * FROM ducknng_run_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'CREATE TABLE IF NOT EXISTS client_side_demo(i INTEGER)');
SELECT * FROM ducknng_run_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'INSERT INTO client_side_demo VALUES (10), (11)');

-- RPC helper: fetch row results through the unary query path.
SELECT * FROM ducknng_query_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'SELECT i, i > 10 AS gt_10 FROM client_side_demo ORDER BY i');

-- Primitive transport layer: open a socket handle, dial it, and inspect the registry.
SELECT ducknng_open_socket('req');
SELECT ducknng_dial_socket(1, 'ipc:///tmp/ducknng_sql_client_demo.ipc', 1000);
SELECT * FROM ducknng_list_sockets();

-- Primitive transport layer: send a raw manifest frame and inspect the reply prefix.
SELECT * FROM ducknng_request_socket(1::UBIGINT, from_hex('01000000000000000000000000000000000000000000'), 1000);
SELECT * FROM ducknng_request('ipc:///tmp/ducknng_sql_client_demo.ipc', from_hex('01000000000000000000000000000000000000000000'), 1000);

-- Raw scalar forms now mean raw reply frames as BLOBs.
SELECT substr(hex(ducknng_request_socket_raw(1, from_hex('01000000000000000000000000000000000000000000'), 1000)), 1, 28);

-- Decode raw manifest and exec replies into envelope fields plus extracted text payload.
SELECT ok, version, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(ducknng_get_rpc_manifest_raw('ipc:///tmp/ducknng_sql_client_demo.ipc'));

SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_run_rpc_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', 'CREATE TABLE IF NOT EXISTS client_side_demo(i INTEGER)')
);

-- The generic raw request helper can be decoded the same way.
SELECT ok, version, type_name, name, payload_text
FROM ducknng_decode_frame(
  ducknng_request_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', from_hex('01000000000000000000000000000000000000000000'), 1000)
);

-- Close the client socket handle and stop the demo server.
SELECT ducknng_close_socket(1);
SELECT ducknng_stop_server('sql_client_demo');
```

### REQ/REP `EXEC` via `nanonext` as an interop example

```{r exec-example}
suppressWarnings(suppressPackageStartupMessages({
  library(DBI)
  library(duckdb)
  library(nanonext)
  library(nanoarrow)
}))

# Little-endian integer encoders for the versioned ducknng RPC envelope.
u32le <- function(x) writeBin(as.integer(x), raw(), size = 4L, endian = "little")
u64le <- function(x) {
  x <- as.double(x)
  c(u32le(x %% 2^32), u32le(floor(x / 2^32)))
}
le_u32_from_raw <- function(x) sum(as.double(as.integer(x)) * 256^(0:3))
le_u64_from_raw <- function(x) le_u32_from_raw(x[1:4]) + 2^32 * le_u32_from_raw(x[5:8])
read_u32le <- function(buf, offset) le_u32_from_raw(buf[offset + 0:3])
read_u64le <- function(buf, offset) le_u64_from_raw(buf[offset + 0:7])

# Serialize a one-row exec request as Arrow IPC using nanoarrow.
encode_ducknng_exec_request <- function(sql, want_result = FALSE) {
  con <- rawConnection(raw(), open = "r+")
  on.exit(close(con))
  nanoarrow::write_nanoarrow(
    data.frame(sql = sql, want_result = want_result),
    con
  )
  payload <- rawConnectionValue(con)
  name <- charToRaw("exec")
  c(
    as.raw(1L),                  # envelope version
    as.raw(1L),                  # message type: call
    u32le(0),                    # flags
    u32le(length(name)),         # method-name length
    u32le(0),                    # error-string length on requests
    u64le(length(payload)),      # payload length
    name,
    payload
  )
}

# Decode a reply envelope and, when present, parse the Arrow IPC payload.
decode_ducknng_exec_reply <- function(buf) {
  name_len <- read_u32le(buf, 7)
  error_len <- read_u32le(buf, 11)
  payload_len <- read_u64le(buf, 15)
  name_start <- 23L
  error_start <- name_start + name_len
  payload_start <- error_start + error_len
  payload_end <- payload_start + payload_len - 1L
  payload <- if (payload_len > 0) buf[payload_start:payload_end] else raw()
  list(
    version = as.integer(buf[1]),
    type = as.integer(buf[2]),
    flags = read_u32le(buf, 3),
    method = rawToChar(buf[name_start:(error_start - 1L)]),
    error = if (error_len > 0) rawToChar(buf[error_start:(payload_start - 1L)]) else "",
    payload = payload,
    data = if (payload_len > 0) as.data.frame(nanoarrow::read_nanoarrow(payload)) else NULL
  )
}

# Use a temporary IPC path and launch a DuckDB-backed ducknng server in a child process.
ext_path <- normalizePath("build/release/ducknng.duckdb_extension")
ipc_path <- tempfile(pattern = "ducknng_readme_exec_", tmpdir = "/tmp", fileext = ".ipc")
ipc_url <- paste0("ipc://", ipc_path)

server_job <- parallel::mcparallel({
  drv <- duckdb::duckdb(config = list(allow_unsigned_extensions = "true"))
  con <- DBI::dbConnect(drv, dbdir = ":memory:")
  DBI::dbExecute(con, sprintf("LOAD '%s'", ext_path))
  DBI::dbGetQuery(con, sprintf(
    "SELECT ducknng_start_server('sql_exec', '%s', 1, 134217728, 300000, NULL, NULL, NULL)",
    ipc_url
  ))
  Sys.sleep(2)
  rows <- DBI::dbGetQuery(con, "SELECT * FROM ducknng_exec_demo ORDER BY i")
  DBI::dbGetQuery(con, "SELECT ducknng_stop_server('sql_exec')")
  DBI::dbDisconnect(con, shutdown = TRUE)
  rows
})

# Give the listener a moment to come up, then dial it with a normal req socket.
Sys.sleep(1)
req <- nanonext::socket("req", dial = ipc_url, autostart = NA)

# Create a table remotely and inspect the reply metadata.
nanonext::send(req, encode_ducknng_exec_request("CREATE TABLE ducknng_exec_demo(i INTEGER)"), mode = "raw", block = 1000L)
create_reply <- decode_ducknng_exec_reply(nanonext::recv(req, mode = "raw", block = 1000L))
create_reply$data

# Insert rows remotely and inspect the second reply.
nanonext::send(req, encode_ducknng_exec_request("INSERT INTO ducknng_exec_demo VALUES (42), (99)"), mode = "raw", block = 1000L)
insert_reply <- decode_ducknng_exec_reply(nanonext::recv(req, mode = "raw", block = 1000L))
insert_reply$data

# Request result rows directly with want_result = TRUE.
nanonext::send(
  req,
  encode_ducknng_exec_request("SELECT i, i > 50 AS gt_50 FROM ducknng_exec_demo ORDER BY i", want_result = TRUE),
  mode = "raw",
  block = 1000L
)
select_reply <- decode_ducknng_exec_reply(nanonext::recv(req, mode = "raw", block = 1000L))
select_reply$data

# Collect the server-side query result from the child process.
server_rows <- parallel::mccollect(server_job)[[1]]
server_rows

close(req)
```

