---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::knit_engines$set(sql = function(options) {
  code <- paste(options$code, collapse = "\n")
  extension_path <- normalizePath("build/release/ducknng.duckdb_extension", mustWork = FALSE)
  extension_path <- gsub("'", "''", extension_path, fixed = TRUE)
  code <- paste(sprintf("LOAD '%s';", extension_path), code, sep = "\n")
  if (isTRUE(options$eval)) {
    py <- file.path(getwd(), "configure", "venv", "bin", "python3")
    py_script <- tempfile(fileext = ".py")
    writeLines(c(
      "import duckdb, sys",
      "sql = sys.stdin.read()",
      "con = duckdb.connect(config={'allow_unsigned_extensions': 'true'})",
      "for stmt in con.extract_statements(sql):",
      "    rel = con.execute(stmt.query)",
      "    if rel.description is None:",
      "        continue",
      "    cols = [d[0] for d in rel.description]",
      "    rows = rel.fetchall()",
      "    print('\\t'.join(cols))",
      "    for row in rows:",
      "        print('\\t'.join('' if v is None else str(v) for v in row))",
      "    print()"
    ), py_script)
    out <- system2(py, py_script, input = code, stdout = TRUE, stderr = TRUE)
    unlink(py_script)
    status <- attr(out, "status")
    if (!is.null(status) && status != 0) stop(paste(out, collapse = "\n"))
    knitr::engine_output(options, code, out)
  } else {
    knitr::engine_output(options, code, NULL)
  }
})
```

# ducknng

`ducknng` is a pure C DuckDB extension for DuckDB + NNG REQ/REP interop.

Its SQL-facing client and server interface is explicitly modeled over [`r-lib/nanonext`](https://github.com/r-lib/nanonext), taking req/rep sockets, dial/listen semantics, context-oriented NNG usage, and a practical messaging surface as the reference point for what should feel natural from DuckDB SQL. Its RPC framing and tabular payload direction are likewise informed by Arrow IPC based RPC work such as [`sounkou-bioinfo/mangoro`](https://github.com/sounkou-bioinfo/mangoro) and related projects, where a thin envelope is kept separate from Arrow IPC request and reply payloads instead of being buried inside one-off method-specific binaries.

`ducknng` now declares itself explicitly as an unstable C API extension targeting a single DuckDB version line. This is deliberate and should not be hidden behind stable-ABI language. The DuckDB project’s own release-cycle documentation classifies unstable API extensions as version-tied extensions that are not binary-compatible across multiple DuckDB versions, and that is the correct model for `ducknng` while it reaches into unstable ABI territory for deeper Arrow/runtime work. The repo should therefore be read as an unstable-ABI extension that targets one DuckDB release line at a time, not as a stable C API extension promising cross-version binary compatibility. In the current local development/test setup that target line is DuckDB `v1.5.2`, and changing DuckDB versions now requires rebuilding and retargeting the extension on purpose.

At the moment, `ducknng` already exposes a low-level transport layer with socket-open, dial, close, and raw request primitives, alongside SQL-visible server lifecycle control and runtime introspection through `ducknng_start_server()`, `ducknng_stop_server()`, and `ducknng_list_servers()`. That transport surface is operation-oriented rather than transport-specific: transport is autodetected from the URL scheme in the same general style as `nanonext`, so the same API can work with `inproc://`, `ipc://`, `tcp://`, and `tls+tcp://` without spawning a separate helper family for each scheme, while TLS material is passed through reusable `tls_config_id` handles instead of inline transport-specific arguments. The current SQL client socket-handle layer is still honest but narrow: `ducknng_open_socket()` only implements `req` today. On the server side it runs one REP socket with one or more REP contexts, and on top of that transport it already supports a working RPC request/reply path with manifest discovery, metadata-oriented execution, and unary row-returning execution over the current raw wire and Arrow IPC model.

The next documented protocol slice is the session query family: `query_open`, `fetch`, `close`, and `cancel`. In the current contract, `query_open` opens one server-owned query session and returns JSON control metadata, `fetch` is the only method that may return streamed row batches, `close` is the normal explicit cleanup path, and `cancel` is best-effort rather than a guarantee of immediate interruption. Until a real owner-identity model is implemented, that family should still be treated as development scaffolding rather than as a production-safe multi-client surface.

The transport-security utility layer now has a runtime model for TLS configuration handles. Those handles can be assembled from file-backed certificate material, from in-memory PEM strings, or from self-signed development certificates generated on demand. The actual `tls+tcp://` transport wiring remains isolated behind `ducknng_nng_compat.c`, but it is now enabled for real loopback listener and one-shot client-request paths so the public API no longer needs a separate transport-specific helper family. This follows the same broad direction as `nanonext`, which exposes TLS helper facilities backed by mbedTLS rather than forcing every client to hand-roll certificate setup outside the transport layer.

## Functions

```{r function-catalog, echo=FALSE, results='asis'}
knitr::asis_output(paste(readLines("function_catalog/functions.md"), collapse = "\n"))
```

## Build

```sh
make configure
make release
```

See also `NEWS.md` for the current implementation status and planned next steps, and `docs/protocol.md`, `docs/manifest.md`, `docs/security.md`, `docs/registry.md`, and `docs/types.md` for the binding transport, TLS, and session/query-family contract. Those design docs also define the current TLS direction: certificate and key material can be accepted either from files or from in-memory PEM content, with helper utilities for development certificate generation and client/server TLS configuration rather than a file-only setup path.

## Examples

### Start an IPC listener and inspect the registry

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'sql0',                         -- service name
  'ipc:///tmp/ducknng_sql0.ipc', -- listen URL
  1,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

SELECT name, listen, contexts, running, sessions
FROM ducknng_list_servers();

SELECT ducknng_stop_server('sql0');
```

### Request multiple REP contexts on one REP socket

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'sql_multi',                    -- service name
  'ipc:///tmp/ducknng_sql_multi.ipc', -- listen URL
  3,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

SELECT name, contexts, running
FROM ducknng_list_servers()
WHERE name = 'sql_multi';

SELECT ducknng_stop_server('sql_multi');
```

### DuckDB can also act as a client

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Start a local ducknng service that the following client examples will talk to.
SELECT ducknng_start_server(
  'sql_client_demo',              -- service name
  'ipc:///tmp/ducknng_sql_client_demo.ipc', -- listen URL
  1,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

-- The default RPC surface is minimal: manifest is built in, exec is opt-in.
SELECT * FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT);
SELECT * FROM ducknng_list_methods();

-- Register exec explicitly before exposing SQL execution over RPC.
SELECT ducknng_register_exec_method();
SELECT * FROM ducknng_list_methods();
SELECT * FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT);

-- RPC helper: run non-row statements and keep errors in-band.
SELECT * FROM ducknng_run_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'CREATE TABLE IF NOT EXISTS client_side_demo(i INTEGER)', 0::UBIGINT);
SELECT * FROM ducknng_run_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'INSERT INTO client_side_demo VALUES (10), (11)', 0::UBIGINT);

-- RPC helper: fetch row results through the unary query path.
SELECT * FROM ducknng_query_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'SELECT i, i > 10 AS gt_10 FROM client_side_demo ORDER BY i', 0::UBIGINT);

-- Primitive transport layer: open a socket handle, dial it, and inspect the registry.
SELECT ducknng_open_socket('req');
SELECT ducknng_dial_socket(1, 'ipc:///tmp/ducknng_sql_client_demo.ipc', 1000);
SELECT * FROM ducknng_list_sockets();

-- Primitive transport layer: send a raw manifest frame and inspect the reply prefix.
SELECT * FROM ducknng_request_socket(1::UBIGINT, from_hex('01000000000000000000000000000000000000000000'), 1000);
SELECT * FROM ducknng_request('ipc:///tmp/ducknng_sql_client_demo.ipc', from_hex('01000000000000000000000000000000000000000000'), 1000, 0::UBIGINT);

-- Raw scalar forms now mean raw reply frames as BLOBs.
SELECT substr(hex(ducknng_request_socket_raw(1, from_hex('01000000000000000000000000000000000000000000'), 1000)), 1, 28);

-- Decode raw manifest and exec replies into envelope fields plus extracted text payload.
SELECT ok, version, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(ducknng_get_rpc_manifest_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT));

SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_run_rpc_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', 'CREATE TABLE IF NOT EXISTS client_side_demo(i INTEGER)', 0::UBIGINT)
);

-- The generic raw request helper can be decoded the same way.
SELECT ok, version, type_name, name, payload_text
FROM ducknng_decode_frame(
  ducknng_request_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', from_hex('01000000000000000000000000000000000000000000'), 1000, 0::UBIGINT)
);

-- Close the client socket handle and stop the demo server.
SELECT ducknng_close_socket(1);
SELECT ducknng_stop_server('sql_client_demo');
```

### `tls+tcp://` with a self-signed development TLS config

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Generate a self-signed loopback certificate and keep it as a runtime TLS handle.
SELECT ducknng_self_signed_tls_config('127.0.0.1', 365, 0);

-- Start a TLS listener using the generated config handle.
SELECT ducknng_start_server(
  'tls_demo_self',
  'tls+tcp://127.0.0.1:45453',
  1,
  134217728,
  300000,
  1
);

-- Send a raw manifest request over TLS and decode the reply frame.
SELECT ok, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'tls+tcp://127.0.0.1:45453',
    from_hex('01000000000000000000000000000000000000000000'),
    1000,
    1
  )
);

-- Clean up the TLS demo server and config handle.
SELECT ducknng_stop_server('tls_demo_self');
SELECT ducknng_drop_tls_config(1);
```

### `tls+tcp://` from file-backed certificate material

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Register a file-backed TLS config using committed loopback test certificates.
SELECT ducknng_tls_config_from_files(
  'test/certs/loopback-cert-key.pem',
  'test/certs/loopback-ca.pem',
  NULL,
  0
);

-- Start a TLS listener with that file-backed config.
SELECT ducknng_start_server(
  'tls_demo_files',
  'tls+tcp://127.0.0.1:45454',
  1,
  134217728,
  300000,
  1
);

-- Inspect the manifest reply over TLS using the same config handle on the client side.
SELECT ok, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'tls+tcp://127.0.0.1:45454',
    from_hex('01000000000000000000000000000000000000000000'),
    1000,
    1
  )
);

-- Clean up the file-backed TLS demo server and config handle.
SELECT ducknng_stop_server('tls_demo_files');
SELECT ducknng_drop_tls_config(1);
```

### REQ/REP `EXEC` via `nanonext` as an interop example

```{r exec-example}
# Little-endian helpers for the versioned ducknng RPC envelope.
write_u32le <- function(x) {
  writeBin(as.integer(x), raw(), size = 4L, endian = "little")
}

write_u64le <- function(x) {
  x <- as.double(x)
  c(write_u32le(x %% 2^32), write_u32le(floor(x / 2^32)))
}

read_u32le_bytes <- function(x) {
  sum(as.double(as.integer(x)) * 256^(0:3))
}

read_u64le_bytes <- function(x) {
  read_u32le_bytes(x[1:4]) + 2^32 * read_u32le_bytes(x[5:8])
}

read_u32le <- function(buf, offset) {
  read_u32le_bytes(buf[offset + 0:3])
}

read_u64le <- function(buf, offset) {
  read_u64le_bytes(buf[offset + 0:7])
}

# Encode the Arrow IPC payload carried by the built-in exec RPC method.
encode_ducknng_exec_payload <- function(sql, want_result = FALSE) {
  raw_con <- rawConnection(raw(), open = "r+")
  on.exit(close(raw_con), add = TRUE)

  nanoarrow::write_nanoarrow(
    data.frame(sql = sql, want_result = want_result),
    raw_con
  )

  rawConnectionValue(raw_con)
}

# Build a version 1 RPC call frame for the named method.
encode_ducknng_call_frame <- function(method, payload, flags = 0L) {
  method_name <- charToRaw(method)

  c(
    as.raw(1L),
    as.raw(1L),
    write_u32le(flags),
    write_u32le(length(method_name)),
    write_u32le(0),
    write_u64le(length(payload)),
    method_name,
    payload
  )
}

encode_ducknng_exec_request <- function(sql, want_result = FALSE) {
  encode_ducknng_call_frame(
    method = "exec",
    payload = encode_ducknng_exec_payload(sql, want_result = want_result)
  )
}

# Decode a reply envelope and parse its Arrow IPC payload when present.
decode_ducknng_exec_reply <- function(buf) {
  header_size <- 23L

  if (!is.raw(buf) || length(buf) < (header_size - 1L)) {
    stop("ducknng reply frame was empty or truncated", call. = FALSE)
  }

  name_len <- read_u32le(buf, 7)
  error_len <- read_u32le(buf, 11)
  payload_len <- read_u64le(buf, 15)

  name_start <- header_size
  error_start <- name_start + name_len
  payload_start <- error_start + error_len
  payload_end <- payload_start + payload_len - 1L

  payload <- if (payload_len > 0) buf[payload_start:payload_end] else raw()

  list(
    version = as.integer(buf[1]),
    type = as.integer(buf[2]),
    flags = read_u32le(buf, 3),
    method = rawToChar(buf[name_start:(error_start - 1L)]),
    error = if (error_len > 0) rawToChar(buf[error_start:(payload_start - 1L)]) else "",
    payload = payload,
    data = if (payload_len > 0) as.data.frame(nanoarrow::read_nanoarrow(payload)) else NULL
  )
}

# Send and receive with the same retry style used in mangoro examples.
send_ducknng_frame <- function(socket, frame, timeout_ms = 1000L, max_attempts = 20L) {
  attempt <- 1L
  send_result <- nanonext::send(socket, frame, mode = "raw", block = timeout_ms)

  while (nanonext::is_error_value(send_result) && attempt < max_attempts) {
    Sys.sleep(0.25)
    send_result <- nanonext::send(socket, frame, mode = "raw", block = timeout_ms)
    attempt <- attempt + 1L
  }

  send_result
}

recv_ducknng_frame <- function(socket, timeout_ms = 1000L, max_attempts = 20L) {
  attempt <- 1L
  response <- nanonext::recv(socket, mode = "raw", block = timeout_ms)

  while (
    (
      nanonext::is_error_value(response) ||
      !is.raw(response) ||
      length(response) < 22L
    ) &&
    attempt < max_attempts
  ) {
    Sys.sleep(0.25)
    response <- nanonext::recv(socket, mode = "raw", block = timeout_ms)
    attempt <- attempt + 1L
  }

  response
}

run_ducknng_req <- function(socket, frame, timeout_ms = 1000L) {
  send_ducknng_frame(socket, frame, timeout_ms = timeout_ms)
  decode_ducknng_exec_reply(
    recv_ducknng_frame(socket, timeout_ms = timeout_ms)
  )
}

# Wait for an IPC listener path to appear before dialing it.
wait_for_ducknng_listener <- function(path, timeout_secs = 10, interval_secs = 0.1) {
  deadline <- Sys.time() + timeout_secs

  while (Sys.time() < deadline) {
    if (file.exists(path)) {
      return(invisible(TRUE))
    }
    Sys.sleep(interval_secs)
  }

  stop("ducknng IPC listener did not become ready in time", call. = FALSE)
}

# Start a ducknng server in this R session, then talk to it over a req socket.
ext_path <- normalizePath("build/release/ducknng.duckdb_extension")
ipc_path <- tempfile(pattern = "ducknng_readme_exec_", tmpdir = "/tmp", fileext = ".ipc")
ipc_url <- paste0("ipc://", ipc_path)

db_driver <- duckdb::duckdb(
  dbdir = ":memory:",
  config = list(allow_unsigned_extensions = "true")
)
db_con <- DBI::dbConnect(db_driver)

DBI::dbGetQuery(db_con, "SELECT 42 AS ok")
DBI::dbExecute(db_con, sprintf("LOAD '%s'", ext_path))
DBI::dbGetQuery(
  db_con,
  sprintf(
    "SELECT ducknng_start_server('sql_exec', '%s', 1, 134217728, 300000, 0::UBIGINT)",
    ipc_url
  )
)
DBI::dbGetQuery(db_con, "SELECT ducknng_register_exec_method()")

wait_for_ducknng_listener(ipc_path)
req <- nanonext::socket("req", dial = ipc_url)

# Create a table remotely and inspect the metadata reply.
create_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request("CREATE TABLE ducknng_exec_demo(i INTEGER)")
)
create_reply$data

# Insert rows remotely and inspect the second metadata reply.
insert_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request("INSERT INTO ducknng_exec_demo VALUES (42), (99)")
)
insert_reply$data

# Request result rows directly with want_result = TRUE.
select_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request(
    "SELECT i, i > 50 AS gt_50 FROM ducknng_exec_demo ORDER BY i",
    want_result = TRUE
  )
)
select_reply$data

# Inspect the same table locally through the DuckDB connection that owns the server.
server_rows <- DBI::dbGetQuery(db_con, "SELECT * FROM ducknng_exec_demo ORDER BY i")
server_rows

# Stop the server and clean up the socket and temporary IPC path.
DBI::dbGetQuery(db_con, "SELECT ducknng_stop_server('sql_exec')")
close(req)
DBI::dbDisconnect(db_con)
unlink(ipc_path)
```

